The Oumupo, between music and math

### 2017-05-03 article.tex
typesets program

\documentclass{article}
\title{The Oumupo, between music and math}
\author{M Andreatta, M Granger, T Johnson and V Villenave
(tr M Solomon)}
\begin{document}
\maketitle
\begin{abstract}
What if music was, at the end of the day, an accessible, fun and expressive way to engage with math? The musicians of the OuMuPo (Ouvroir de Musique Potentielle, or The Workshop for Potential Music) lead us through this open question via different exercises and experiments.
\end{abstract}

@@

\begin{itemize}
\item \textbf{rhythms} including the duration of notes and rests, but as
well the number of repetitions of a note or a group of notes.  This even
extends to the quantity of events within a single structural unit (for
example, the number of notes in a measure).
\item \textbf{absolute pitch}, or the degrees of the diatonic scale that are
often numbered with Roman Numerals or designated with letters of the
alphabet. Since the twentieth century, it is even possible to designate a
pitch by its periodic frequency in Hz!
\item \textbf{relative pitch}, or the interval between a pitch and another
one, played either disjointly or simultaneously in time. By counting the
number of semitones between two notes, it is possible to establish detailed
metrics of chords and harmonies. The \emph{Tonnetz} (see below) is, in this
line of thought, an original and interesting tool.
\end{itemize}
In many cases, the transation of a number or a mathematical operation
into musical elements.  This leads to the questioning of the \textbf{numeric
base} being used.  While we are used to manipulating numbers in a base 10
scheme, musical organization can be conceived in entirely different ways.
\begin{itemize}
\item Beats and measures are often counted in units that are $2^x$, leading
to a measurement system in base 4 (the bread-and-butter of European Common Practice
musical phrasing) or 8 (the ``count to eight'' common for dancers).
\item The diatonic scale, an omnipresent seven-note construct in many different musical
traditions including the Western one. The pentatonic scale, which comprises
five notes, is found in many Eastern cultures.
\item The chromatic scale, used since the time of Pythagorian theorists,
requires the partitioning of a musical octave in a twelve-semitone scale.
Today, this scale is a strict geometric series (each note incrementing by a
factor of $2^{1/12}$).
\end{itemize}
It is nevertheless possible to cheat through various means of musical
cunning.  For example, a ten-note set can be derived from the twelve-note
integrality by omitting two notes from the chromatic collection.
\section*{Melodic transformation}
Forms such as the canon have existed for several centuries.  Numerous
composers, including the well-known J.S. Bach, have proven that it is
possible to make musical counterpoint by playing the same melody
simultaneously at different tempos.  With this in mind, we attempt to
explore new melodic and polyphonic structures.  One example of this is
self-similar melodies, meaning melodies that reproduce themselves at
different scales.  In mathematics, this idea is encapsulated by fractals. 
Concretely, the melody rearticulates itself by only playing one note out of
every \emph{n} notes.  In the most complex cases, \emph{n} can change but
the melody remains invariant.
\section*{Combinatorial procedures}
Music, compared to literature, is combinatorially agnostic, meaning that
the constraint of meaning attributed to words that make certain literary
combinations valid and others not is more difficult to articulate in music. 
For example, the French word \emph{mode} can have its letters reorganized
into \emph{d\^{o}me} or \emph{d\'{e}mo} whereas \emph{doem} or \emph{medo}
are not words in the French lexicon.  Musical notes, on the other hand, can
be combined in any order and carry a subjective argument in musical time.
Several classic mathematic tools can be used to articulate musics
combinatorial potential: magic squares, Pascal's triangle, the Dragon's
Curve (see below) and many others.
\end{document}

>>>>

\documentclass{article}
\usepackage{listings}
\title{The Oumupo, between music and math}
\author{M Andreatta, M Granger, T Johnson and V Villenave
(tr M Solomon)}
\begin{document}
\maketitle
\begin{abstract}
What if music was, at the end of the day, an accessible, fun and expressive way to engage with math? The musicians of the OuMuPo (Ouvroir de Musique Potentielle, or The Workshop for Potential Music) lead us through this open question via different exercises and experiments.
\end{abstract}

@@

\begin{itemize}
\item \textbf{rhythms} including the duration of notes and rests, but as
well the number of repetitions of a note or a group of notes.  This even
extends to the quantity of events within a single structural unit (for
example, the number of notes in a measure).
\item \textbf{absolute pitch}, or the degrees of the diatonic scale that are
often numbered with Roman Numerals or designated with letters of the
alphabet. Since the twentieth century, it is even possible to designate a
pitch by its periodic frequency in Hz!
\item \textbf{relative pitch}, or the interval between a pitch and another
one, played either disjointly or simultaneously in time. By counting the
number of semitones between two notes, it is possible to establish detailed
metrics of chords and harmonies. The \emph{Tonnetz} (see below) is, in this
line of thought, an original and interesting tool.
\end{itemize}
In many cases, the transation of a number or a mathematical operation
into musical elements.  This leads to the questioning of the \textbf{numeric
base} being used.  While we are used to manipulating numbers in a base 10
scheme, musical organization can be conceived in entirely different ways.
\begin{itemize}
\item Beats and measures are often counted in units that are $2^x$, leading
to a measurement system in base 4 (the bread-and-butter of European Common Practice
musical phrasing) or 8 (the ``count to eight'' common for dancers).
\item The diatonic scale, an omnipresent seven-note construct in many different musical
traditions including the Western one. The pentatonic scale, which comprises
five notes, is found in many Eastern cultures.
\item The chromatic scale, used since the time of Pythagorian theorists,
requires the partitioning of a musical octave in a twelve-semitone scale.
Today, this scale is a strict geometric series (each note incrementing by a
factor of $2^{1/12}$).
\end{itemize}
It is nevertheless possible to cheat through various means of musical
cunning.  For example, a ten-note set can be derived from the twelve-note
integrality by omitting two notes from the chromatic collection.
\section*{Melodic transformation}
Forms such as the canon have existed for several centuries.  Numerous
composers, including the well-known J.S. Bach, have proven that it is
possible to make musical counterpoint by playing the same melody
simultaneously at different tempos.  With this in mind, we attempt to
explore new melodic and polyphonic structures.  One example of this is
self-similar melodies, meaning melodies that reproduce themselves at
different scales.  In mathematics, this idea is encapsulated by fractals. 
Concretely, the melody rearticulates itself by only playing one note out of
every \emph{n} notes.  In the most complex cases, \emph{n} can change but
the melody remains invariant.
\section*{Combinatorial procedures}
Music, compared to literature, is combinatorially agnostic, meaning that
the constraint of meaning attributed to words that make certain literary
combinations valid and others not is more difficult to articulate in music. 
For example, the French word \emph{mode} can have its letters reorganized
into \emph{d\^{o}me} or \emph{d\'{e}mo} whereas \emph{doem} or \emph{medo}
are not words in the French lexicon.  Musical notes, on the other hand, can
be combined in any order and carry a subjective argument in musical time.
Several classic mathematic tools can be used to articulate musics
combinatorial potential: magic squares, Pascal's triangle, the Dragon's
Curve (see below) and many others.
\section*{Attractors}
Attractors in music are a tool used by composers looking for a music that
contains qualities of both structure and chaos.

One interesting example of attractors in music comes from the Moscow-based
composer Sergei Zagny (born in 1960) and is based on mathematical equations
elaborated by the Belgian mathemetician Pierre-Fran\c{c}ois Verhulst
(1804-1849). A variable $x$ (which must always stay between 0 and 1) is
recursively multiplied by a factor $a$ (that can be between 0 and 4) and we
observe the following iterative results.

\begin{lstlisting}
a = 2.6; x = 0.5; ct = 0;
While[ct <= 30, ct++; x = N[a*x*(1 - x)];
Print[ct, "  ", x]]

0.652  0.59153  0.6282324  0.6072475  0.6200956  0.6125017
0.6170938  0.6143529 0.61600210  0.61501311  0.61560712 
0.61525113  0.61546514  0.61533715  0.61541316 0.61536717
0.61539518  0.61537819  0.61538820  0.61538221  0.61538622
0.61538423 0.61538524  0.61538425  0.61538526  0.61538527
0.61538528  0.61538529  0.61538530 0.61538531  0.615385
\end{lstlisting}
\end{document}
